---
title: "Statistical report: Loop Diuretic Therapy as a Prognostic Enrichment Factor for Clinical Trials in Patients with Heart Failure with mrEF and pEF"
date: "`r Sys.Date()`"
cover: static/ki_logo_vert_rgb.png
editor: source
format:
  pdf:
    documentclass: scrartcl
    template-partials:
      - "static/before-body.tex"
      - "static/_titlepage.tex"
    include-in-header: 
      - "static/in-header.tex"
    toc: true
    toc-depth: 3
    toc-title: Contents
    lof: true
    lot: true
    cap-location: top
    number-sections: true
    colorlinks: false
    keep-tex: false
    
bibliography: static/references.bib
csl: static/bmj.csl
link-citations: true
link-bibliography: true
nocite: '@*'
knitr:
  opts_chunk: 
    dev: cairo_pdf
    collapse: true
    comment: "" 
    echo: FALSE
    include: TRUE
    warning: FALSE
    message: FALSE
    fig.pos: "H"
    fig.path: "../output/figs/"
    fig.height: 6
    fig.width: 10
    R.options:
      knitr.graphics.auto_pdf: true
---

```{r}
#| label: set-up-load-data
options(knitr.kable.NA = "")

# load packages, globals and project specific functions
source(here::here("setup/setup.R"))

# load data
load(here("data/clean-data/data.RData"))

# load workbook to write tables to Excel
wb <- loadWorkbook(here("output/tabs/tables.xlsx"))
sheets <- names(wb)
```           

\newpage

# Data

## Data sources

The Swedish Heart Failure Registry (SwedeHF) is a nationwide quality registry started in 2000, 
that includes in- and out-hospital patients with HF, regardless of EF and previous history of HF[@swedehf]. 
Around 80 variables are recorded at 
discharge from hospital or at the outpatient visit, i.e. index date, and 
entered into an electronic database managed by the Uppsala Clinical Research 
Center (Uppsala, Sweden). Up to April 2017 the only inclusion criterion was a 
clinical diagnosis of HF, which was thereafter defined according to the ICD-10 
codes I50.0, I50.1, I50.9, I42.0, I42.6, I42.7, I25.5, I11.0, I13.0, I13.2. 
Informed consent is not required, although patients are informed of entry and can opt out. 
In 2021, 69 of 76 Swedish hospitals enrolled patients in SwedeHF, which had a 32% coverage of the prevalent HF population in Sweden[@annualreport]. 

SwedeHF was linked with national Swedish registries (SHFDB v 4.1.0) through the 
Swedish personal identity number (PIN)[@pin]: 

- The National Patient Register (The National Board of Health and Welfare) for 
additional comorbidities and the hospitalization outcomes
- The Cause of Death Register (The National Board of Health and Welfare) for outcomes 
death
- LISA[@lisa] and the Register of the Total Population (Statistics Sweden) for 
socio-economic factors

## Acknowledgments

Please include "We thank all staff members at all care units in Sweden for their contribution to the SwedeHF register." 
in the Acknowledgements of any resulting articles. 

## Ethics

The study was approved by the Swedish Ethical Review Authority, dnr 2021-04326. 

\newpage

## Inclusion/exclusion criteria

```{r}
#| label: tbl-flow
#| tbl-cap: Information for flowchart
default_kable(flow) %>%
  row_spec(c(1, 10, 19, 36, 37), bold = T)
```

## Definitions

Information on data sources, definitions etc. are found https://kiheartfailure.github.io/shfdb4/.

{{< include src/vars.qmd >}}

### Loop diuretic

All analyses are performed for No loop diuretic, Dose 1-40mg/24h, Dose >40mg/24h. Doses were converted to Furosemid equivalents were Furosemide 40 mg = Toresamid 20 mg = Bumetanid 1 mg. 
When there was no information on substance it was assumed to be Furosemide. 

Prescribed daily usage of loop diuretics was considered as taking 
loop diuretics and when necessary as not taking loop diuretic. 

Note that loop diuretic usage can be either prevalent use or newly initiated use. 

# Statistical analysis 

## General

All analyses were performed using `r sessionInfo()$R.version$version.string` [@r]. 
The level of significance is set to 5%, two-sided. No adjustment for multiple 
comparisons were made and therefore the results should be viewed with care.

## Missing data

Missing data was imputed with multiple imputation (n = 10) using mice [@mice]. 
Variables included in the model are indicated in @tbl-base. 
All-cause mortality was included as the Nelson-Aalen estimator.

## Populations

All analyses will be performed for both population 1 and population 2 (see @tbl-flow for definitions). 

## Baseline characteristics

Baseline characteristics are presented by loop diuretic
were categorical variables are presented with n and percentage and tested for 
differences between groups with the chi-square test. 
Continuous variables are presented with median [first quartile-third quartile] and tested for 
differences between groups with the Kruskal-Wallis test. 

## Association between loop diuretic and outcomes

The following outcomes are considered: 

- CV death (CVD) + first HF hospitalization (HFH)
- CVD + total HFH
- CVD
- First HFH
- Total HFH
- All-cause death

Time to first events were presented with cumulative incidence curves (CIC) and 
the mean cumulative function (MCF) was used to depict the recurrent events by loop diuretic. 

Crude and adjusted analyses were performed where adjustment was performed for 
variables indicated in @tbl-base. 
The variables were selected based on clinical relevance.

Cox proportional hazards regressions were used to model the time to first event. 
As a consistency analysis the time to first events were modelled using a 
sub-distributional hazards model[@fg] 
where death was treated as a competing event (only crude models). 

Recurrent events were model using a negative binomial regression including the 
log of time as an offset in the model. 

Data were censored at `r global_followup/365.25` years follow-up, 2021-12-31, 
emigration from Sweden or, when it was itself not an event, death. 

### Assumptions

The proportional hazards assumption was investigated using the scaled Schoenfeld 
residuals for the primary outcome leading to that location was stratified for in the models. 
Possible outliers were visually inspected by plotting the dfbetas. 

# Results

```{r}
med_pop1 <- rsdata %>%
  filter(pop1) %>%
  summarise(
    med = fn(median(shf_age), dig = 1),
    q1 = fn(quantile(shf_age, probs = 0.25), dig = 1),
    q3 = fn(quantile(shf_age, probs = 0.75), dig = 1)
  ) %>%
  mutate(out = paste0(med, " (", q1, "-", q3, ")")) %>%
  pull(out)

percfem_pop1 <- rsdata %>%
  filter(pop1) %>%
  count(shf_sex) %>%
  mutate(perc = fn(n / sum(n) * 100, 1)) %>%
  filter(shf_sex == "Female") %>%
  pull(perc)

med_pop2 <- rsdata %>%
  summarise(
    med = fn(median(shf_age), dig = 1),
    q1 = fn(quantile(shf_age, probs = 0.25), dig = 1),
    q3 = fn(quantile(shf_age, probs = 0.75), dig = 1)
  ) %>%
  mutate(out = paste0(med, " (", q1, "-", q3, ")")) %>%
  pull(out)

percfem_pop2 <- rsdata %>%
  count(shf_sex) %>%
  mutate(perc = fn(n / sum(n) * 100, 1)) %>%
  filter(shf_sex == "Female") %>%
  pull(perc)
```

The median (q1-q3) age is `r med_pop1` and `r percfem_pop1`% females in Population 1 and 
`r med_pop2` and `r percfem_pop2`% females in Population 2.    

{{< include src/dose_barplot.qmd >}}

\blandscape

## Baseline characteristics

{{< include src/base_tab.qmd >}}
{{< include src/base_tab_af.qmd >}}

\elandscape

## Association between loop diuretic and outcomes

```{r}
medsurv_pop1 <- rsdata %>%
  filter(pop1) %>%
  summarise(
    med = fn(median(sos_outtime_death / 365.25 * 12), dig = 1),
    min = fn(min(sos_outtime_death / 365.25 * 12), dig = 1),
    max = fn(max(sos_outtime_death / 365.25 * 12), dig = 1)
  ) %>%
  mutate(out = paste0(med, " (", min, "-", max, ")")) %>%
  pull(out)

sumsurv_pop1 <- rsdata %>%
  filter(pop1) %>%
  summarise(sumpy = fn(sum(sos_outtime_death) / 365.25, dig = 0)) %>%
  pull(sumpy)

medsurv_pop2 <- rsdata %>%
  summarise(
    med = fn(median(sos_outtime_death / 365.25 * 12), dig = 1),
    min = fn(min(sos_outtime_death / 365.25 * 12), dig = 1),
    max = fn(max(sos_outtime_death / 365.25 * 12), dig = 1)
  ) %>%
  mutate(out = paste0(med, " (", min, "-", max, ")")) %>%
  pull(out)

sumsurv_pop2 <- rsdata %>%
  summarise(sumpy = fn(sum(sos_outtime_death) / 365.25, dig = 0)) %>%
  pull(sumpy)
```

The median (min-max) follow-up is 
`r medsurv_pop1` months for a total of `r sumsurv_pop1` patient-years of follow-up in Population 1 and 
`r medsurv_pop2` months for a total of `r sumsurv_pop2` patient-years of follow-up in Population 2.

{{< include src/outcomes.qmd >}}

{{< include src/km_figs.qmd >}}

\clearpage

# Reproducibility

## R session information {#sessioninfo}

```{r sessinfo}
sessionInfo()
```

## R code

The R code for all data handling and statistical analyses are found: 

https://github.com/KIHeartFailure/swedehf-loopdiuretic. On publication
the repository will be made public so as to 
link to it from the resulting article for increased transparency and code sharing.
No data or output is stored in the repository. 

# References

::: {#refs}
:::
